---
title: "R Project - Hugues Rubin de Cervens"
output:
  html_document:
    df_print: paged
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## **Introduction**

The aim of this report is to build a portfolio of equities and test two strategies to beat the CAC40 index, over the period from January 1, 2018 to December 31, 2023.
The portfolio in question here contains 8 stocks of French companies in various sectors:

-TotalEnergies SE (TTE.PA), energy sector 

-L'Oréal S.A (OR.PA), beauty and cosmetics industry

-Sanofi (SAN.PA), pharmaceutical sector 

-BNP Paribas (BNP.PA), banking sector 

-Airbus SE (AIR.PA), aerospace and defense industry

-Orange S.A (ORA.PA), telecommunications sector

-Vinci S.A (DG.PA), construction and materials sector

-Danone S.A (BN.PA), food sector

### List of libraries used

```{r, message=FALSE, warning=FALSE}
library(quantmod)
library(ggplot2)
library(tseries)
library(dplyr)
library(reshape2)
library(TTR)
library(rmarkdown)
```

## **Part 1** : Equally weighted Portfolio

The first strategy is simply to build our portfolio in such a way that each stock has the same weight (equally-weighted portfolio). 

```{r, echo=TRUE, warning=FALSE}
tickers <- c("TTE.PA", "OR.PA", "SAN.PA", "BNP.PA", "AIR.PA", "ORA.PA", "DG.PA", "BN.PA")
stock_data <- list()

for (ticker in tickers) {
  stock_data[[ticker]] <- getSymbols(ticker, from = "2018-01-01", to = "2023-12-31", src = "yahoo", auto.assign = FALSE)
}

head(stock_data['TTE.PA'])
```
### Calculating portfolio return over time
```{r, echo=TRUE, warning=FALSE}

returns_data <- lapply(stock_data, function(x) dailyReturn(Cl(x)))
portfolio_returns <- do.call(cbind, returns_data) %>% rowMeans(na.rm = TRUE)
portfolio_returns_ts <- xts(portfolio_returns, order.by = index(returns_data[[1]]))
cumulative_returns <- cumprod(1 + portfolio_returns_ts)
```

### Plotting the return
```{r, echo=TRUE, warning=FALSE}

ggplot(data = fortify(cumulative_returns), aes(x = Index, y = coredata(cumulative_returns))) +
  geom_line() +
  ggtitle("Evolution of the equally weighted portfolio return (2010-2023)") +
  xlab("Date") +
  ylab("Cumulative value of the portfolio") +
  theme_minimal()
```

Our second objective is to compare the performance of our portfolio with the CAC40 index. 

### Plotting CAC40 return

```{r, echo=TRUE, warning=FALSE}

cac40_ticker <- "^FCHI"
cac40_data <- getSymbols(cac40_ticker, from = "2018-01-01", to = "2023-12-31", src = "yahoo", auto.assign = FALSE, na.omit = TRUE)
cac40_data_xts <- xts(cac40_data$FCHI.Close, order.by = index(cac40_data))
cac40_returns <- dailyReturn(cac40_data_xts, type = 'log')  # type = 'log' ou type = 'arithmetic'
cumulative_returns <- cumprod(1 + cac40_returns)
cumulative_returns_fortified <- fortify(cumulative_returns)

ggplot(data = cumulative_returns_fortified, aes(x = Index, y = coredata(cumulative_returns))) +
  geom_line(color = 'red') +
  ggtitle("Return on the CAC40 Index (2018-2023)") +
  xlab("Date") +
  ylab("cumulated return") +
  theme_minimal()
```

### Results of the equally-weighted portfolio VS CAC40 index

```{r, echo=TRUE, warning=FALSE}

# For the Sharpe Ratio, we have to download the risk free rate using the latest T-bills data

getSymbols("DTB3", src = "FRED")

riskfree <- as.numeric(last(DTB3)[,1]) / 100

print(riskfree)

```

```{r, echo=TRUE, warning=FALSE}
# Annual return of the portfolio
annual_return <- mean(portfolio_returns) * 252

# Annual volatility of the portfolio
annual_volatility <- sd(portfolio_returns) * sqrt(252)

# Sharpe ratio of the portfolio
sharpe_ratio <- (mean(portfolio_returns)-riskfree / 252) / sd(portfolio_returns) * sqrt(252)

max_drawdown <- maxdrawdown(cumulative_returns)

# CAC40 total return over the covered period
cac40_returns <- dailyReturn(cac40_data_xts)
cumulative_cac40_returns <- sum(cac40_returns)

# beta of the portfolio
aligned_returns <- na.omit(merge(portfolio_returns_ts, cac40_returns, join = "inner"))
covariance <- cov(aligned_returns[, 1], aligned_returns[, 2])
cac40_variance <- var(aligned_returns[, 2])
portfolio_beta <- covariance / cac40_variance
cumulative_portfolio_return <- sum(portfolio_returns_ts)

# Displaying the results
cat("Mean annual yield of the portfolio:", annual_return, "\n")
cat("Total yield of the portfolio:", cumulative_portfolio_return, "\n")
cat("Total yield of the CAC40:", cumulative_cac40_returns, "\n")
cat("Portfolio beta:", portfolio_beta, "\n")
cat("Annual volatility of the portfolio:", annual_volatility, "\n")
cat("Sharpe ratio of the portfolio:", sharpe_ratio, "\n")
```

As we can see above, the beta of my portfolio is relatively close to 1 (0.92) which makes sense, since the values I chose to include in my portfolio are stocks listed on the french stocks market and included in the CAC40 index. Therefore, my portfolio is highly correlated with the index.

However, we can observe that the total return of the portfolio (46%) is slightly below the return of the CAC40 during the same period (47%).

Finally, the sharpe ratio of my portfolio is relatively low (0.11) which means that the portfolio generates low excess return per unit of risk taken.

## **Part 2** : Advanced strategy

For the advanced strategy, the objective is to beat the CAC40 index using technical analysis and active trading.
In order to do this, I chose to use two technical analysis tools : the moving averages and the RSI.

- RSI : The RSI (Relative Strength Index) is an indicator measuring the magnitude of recent changes in price of an asset. We consider the asset as overbought if the value of the RSI is beyond 70 and oversold if the value of the RSI is below 30. Therefore, for the strategy, the objective is to buy the stock of the portfolio if its RSI<30 and sell the stock if the RSI>70.

- Moving averages : Moving averages calculates the average price over a set period and allows the investor to identify changes in trends. There is 2 types of moving averages : the simple moving average, also known as SMA, and the exponential moving averages, known as EMA. EMA gives more weight to recent data by reacting quicker to prices changes. However, for my strategy, I decided to use SMA as it is better suited for long-term trend analysis.

### Creation of the functions for the indicators

```{r, echo=TRUE, warning=FALSE}

# Function computing the two moving averages and the RSI. I chose periods of 50 and 200 days as the data is observed on severals years
calculate_indicators <- function(stock_data) {
  rsi <- RSI(Cl(stock_data), n = 14)
  sma_short <- SMA(Cl(stock_data), n = 50)
  sma_long <- SMA(Cl(stock_data), n = 200)
  return(list(rsi = rsi, sma_short = sma_short, sma_long = sma_long))
}

# Function plotting the RSI indicator and the SMA for each stock of my portfolio
plot_stock_data <- function(stock_symbol, stock_data) {
  selected_stock_data <- stock_data[[stock_symbol]]
  indicators <- calculate_indicators(selected_stock_data)
  chartSeries(selected_stock_data, name = stock_symbol, 
              TA = "addSMA(n = 50, col = 'blue'); addSMA(n = 200, col = 'red')")
  plot(indicators$rsi, main = paste("RSI -", stock_symbol))
}

```

### Plotting the indicators for each stock of the portfolio

```{r, echo=TRUE, warning=FALSE, warning=FALSE}
plot_stock_data("BN.PA", stock_data)
plot_stock_data("TTE.PA", stock_data)
plot_stock_data("OR.PA", stock_data)
plot_stock_data("SAN.PA", stock_data)
plot_stock_data("AIR.PA", stock_data)
plot_stock_data("ORA.PA", stock_data)
plot_stock_data("DG.PA", stock_data)
plot_stock_data("BNP.PA", stock_data)
```

Now, the objective of the strategy is to create a new column, called "Signal", for each stock of the portfolio :

for the RSI : 

- If the RSI>70, then the stock has to be shorted. This means that at that moment, the "Signal" column will take the value of -1 and, from that moment, the initial return (the one corresponding to a long position) will be multiplied by -1.

- If the RSI<30, then the "Signal" column will take the value of 1.

For the moving averages :

- When the value of the SMA50 goes under the value of the SMA200, that means that the trend is reversing and that the stock price is falling. Therefore, the "Signal" column will take the value of "-1" to simulate a short position.

- If the value of the SMA50 is rising above the value of the SMA200, then the stock price is going up and the "Signal" Column will take the value of "1" to simulate a long position.

### Creation of the "Signal" column & calculation of the daily returns using the strategy

```{r, echo=TRUE, warning=FALSE}

for (ticker in tickers) {
  donnees <- getSymbols(ticker, from = "2018-01-01", to = "2023-12-31", src = "yahoo", auto.assign = FALSE)
  donnees$SMA50 <- SMA(Cl(donnees), 50)
  donnees$SMA200 <- SMA(Cl(donnees), 200)
  donnees$RSI <- RSI(Cl(donnees), 14)
  
  # Creating the "Signal" column, taking the values of 1 or -1 depending on the values of the RSI and the SMA
  donnees$Signal <- ifelse(donnees$RSI < 30, 1,
                           ifelse(donnees$RSI > 70, -1,
                                  ifelse(donnees$SMA50 > donnees$SMA200, 1,
                                         ifelse(donnees$SMA50 < donnees$SMA200, -1, NA))))
  donnees$Signal <- na.locf(donnees$Signal, na.rm = FALSE)
  donnees$Signal[is.na(donnees$Signal)] <- 0
  donnees$DailyReturn <- Cl(donnees) / lag(Cl(donnees)) - 1
  donnees$DailyReturnWithStrat <- donnees$Signal * donnees$DailyReturn
  stock_data[[ticker]] <- as.data.frame(donnees)
}

```

From now on, for each stock in the portfolio, we can compare the "classic" daily performance with that of the trading strategy implemented previously.

### Comparing the returns for each stock

```{r, echo=TRUE, warning=FALSE}
Return_strat <- function(stock_data) {
  for (ticker in names(stock_data)) {
    donnees <- stock_data[[ticker]]
    sum_daily_return <- sum(donnees$DailyReturn, na.rm = TRUE)
    sum_daily_strat <- sum(donnees$DailyReturnWithStrat, na.rm = TRUE)
    cat("For", ticker, ":\n")
    cat("Sum of daily returns :", sum_daily_return, "\n")
    cat("Sum of daily returns using the strategy :", sum_daily_strat, "\n\n")
    donnees$date <- index(donnees)
    donnees_melt <- melt(donnees[, c('date', 'DailyReturn', 'DailyReturnWithStrat')], id = 'date')
    p <- ggplot(donnees_melt, aes(x = date, y = value, color = variable)) + 
      geom_line() +
      labs(x = "Days", y = "Return", title = paste("Daily return VS Daily return using trading signals for", ticker)) +
      theme_minimal()
    print(p)
  }
}

Return_strat(stock_data)


```

### Computing the average return of the portfolio and comparison with the CAC40 index

```{r, echo=TRUE, warning=FALSE}

sum_return_strat <- sapply(stock_data, function(data) sum(data$DailyReturnWithStrat, na.rm = TRUE))
weight_2 <- c(0.125, 0.125, 0.125, 0.125, 0.125, 0.125, 0.125, 0.125)
mean_return_strat <- sum(sum_return_strat * weight_2)
cat("Weighted average strategy returns for the portfolio :", mean_return_strat, "\n")
cac40_returns <- dailyReturn(Cl(cac40_data))
aligned_returns <- na.omit(merge(portfolio_returns, cac40_returns, join = "inner"))
regression_model <- lm(aligned_returns[, 1] ~ aligned_returns[, 2])
portfolio_beta <- coef(regression_model)[2]
cat("The beta of the portfolio is :", portfolio_beta, "\n")
cac40_returns <- dailyReturn(Cl(cac40_data))
sum_cac40_returns <- sum(cac40_returns, na.rm = TRUE)
cat("The sum of CAC 40 daily returns is :", sum_cac40_returns, "\n")
```
As expected, the active trading strategy failed to beat the benchmark used (CAC40 index). As can be seen from the graphs and average returns for each stock, the strategy's results are rather disappointing, and well below the classic daily returns.
These poor results can be explained by a number of factors:

- First of all, technical indicators are based on historical data. The use of historical data can therefore lead to a delay in signal generation, and the taking of a position (long or short) when the optimum movement has already passed.

- Secondly, we can imagine that the use of just two technical indicators is not necessarily sufficient or well-suited to the portfolio's chosen stocks. The RSI may be effective in a market with a clear trend, and more effective in intraday trading. Similarly, the SMA can generate false signals in a highly volatile market, which can considerably reduce daily returns.

- Finally, indicator settings may be well optimized at time t, but may be obsolete at time t+1 in the event of changing market conditions. This is what happened in 2020 during the COVID-19 crisis, which greatly amplified market volatility over a certain period.

These results also show that technical analysis alone doesn't necessarily produce good results. When creating an equity portfolio, it is also crucial to analyze the economic situation of individual companies, as well as the macro-economy as a whole.

